---
title: "Sequential covariate-adjusted randomization with SCR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sequential covariate-adjusted randomization with SCR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette shows how to use the **SCR** package, which implements the
sequential covariate-adjusted randomization of

> Yang, H., Qin, Y., Li, Y., and Hu, F. (2024). Sequential covariate-adjusted
> randomization via hierarchically minimizing Mahalanobis distance and
> marginal imbalance. *Biometrics*, 80(2), ujae047.

SCR assigns participants one at a time. Covariate balance, measured by the
modified Mahalanobis distance
$\widetilde M(n) = (\sum_{T_i=1} x_i - \sum_{T_i=0} x_i)^\top \mathrm{cov}(x)^{-1}
(\sum_{T_i=1} x_i - \sum_{T_i=0} x_i)$, is targeted with probability `q1`
only while the group sizes differ by less than `d`. Otherwise the smaller
group is favoured with probability `q2`.

```{r}
library(SCR)
```

## 1. Allocating a sequence of participants

```{r}
set.seed(1)
n <- 500; p <- 10
x <- MASS::mvrnorm(n, mu = rep(0, p), Sigma = 2 * diag(p))

fit <- SCR(x, d = 5, q1 = 0.75, q2 = 0.85)
fit
head(fit$assignment, 20)
```

`fit$prob` gives the probability $P(T_i = 1)$ used for each participant and
`fit$rule` the step of Algorithm 1 that was applied.

```{r}
table(fit$rule)
```

Compared with complete randomization (CR), the Mahalanobis distance under SCR
is close to zero, while under CR it follows a scaled $\chi^2_p$
distribution:

```{r}
M <- replicate(50, {
  x <- matrix(rnorm(n * p), n, p)
  c(SCR = SCR(x)$Mahalanobis_Distance,
    CR = scr_imbalance(x, rbinom(n, 1, 0.5))[["Mahalanobis_Distance"]])
})
rowMeans(M)
```

## 2. Randomizing participants as they arrive

In a trial each participant must be randomized on arrival. `scr_next()`
takes the covariates and assignments of the participants already enrolled
and returns the assignment of the new one.

```{r}
enrolled_x <- x[1:20, ]
enrolled_t <- fit$assignment[1:20]
new_patient <- x[21, ]
scr_next(new_patient, covariate = enrolled_x, assignment = enrolled_t, d = 5)
```

With the same random seed, randomizing the participants one by one with
`scr_next()` gives exactly the same assignments as `SCR()`.

## 3. Continuous and categorical covariates

Categorical covariates enter through the within-stratum imbalance
$\|D_n\|_2^2$, combined with the modified Mahalanobis distance as
$W(n) = w \widetilde M(n) + (1 - w)\|D_n\|_2^2$ (Section 2.2 of the paper).

```{r}
z <- data.frame(sex = sample(c("F", "M"), n, TRUE),
                site = sample(paste0("site", 1:4), n, TRUE))
fit_mixed <- SCR(x, categorical = z, d = 5, w = 0.7)
fit_mixed
```

## 4. Estimating and testing the treatment effect

We simulate outcomes from $y_i = \mu_1 T_i + \mu_2(1 - T_i) + \sum_j \beta_j
x_{ij} + \varepsilon_i$ and compare the difference-in-means estimator with
the regression-adjusted estimator.

```{r}
tr <- fit$assignment
y <- 2 * tr + 1 * (1 - tr) + drop(x %*% rep(1, p)) + rnorm(n)
c(difference_in_means = mean(y[tr == 1]) - mean(y[tr == 0]),
  regression_adjusted = unname(coef(lm(y ~ tr + x))["tr"]))
```

`scr_test()` performs a randomization test: it re-runs SCR on the observed
covariates to obtain the reference distribution of the t statistic.

```{r}
scr_test(y, x, tr, B = 200, d = 5)
```

## 5. More than two treatments

`smart()` extends the procedure to $K$ arms, combining the pairwise modified
Mahalanobis distances by their mean, maximum or median.

```{r}
smart(x, K = 3, d = 5, q = 0.75, method = "mean")
```
